---
title: "CoPheScan: Example with Hierarchical Priors"
author: "Ichcha Manipur"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CoPheScan: Example with Hierarchical Priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "HP-"
)
```

The hierarchical model is ideal when a large set of variants and phenotypes are available i.e., a phenome-wide association study with a set of variants which are previously  known to have a functional role or have been implicated in a disease. 
Here we show how to use cophescan to infer hierarchical priors on a small test dataset.

```{r setup, message=F, warning=F}
library(cophescan)
library(dplyr)
#library(ggpubr)
```

```{r}
data("cophe_multi_trait_data")
attach(cophe_multi_trait_data)
trait_dat = cophe_multi_trait_data$summ_stat$Trait_1
str(trait_dat)
querysnpid <- cophe_multi_trait_data$querysnpid
```


#### Multi-trait analysis

The first step is preparing the input for the hierarchical model which are the log Bayes factors: lBF.Ha and lBF.Hc. We will use `cophe.susie.lbf` to extract Bayes factors estimated using SuSIE.
Note: When there are no credible sets identified with SuSIE the function internally calculates lBF.Ha and lBF.Hc using the Approximate Bayes Factor method. 
```{r message=F, warning=F, results='hide'}
## Hide print messages from coloc
res.multi.lbf <- list()
for (trait_idx in seq_along(cophe_multi_trait_data$summ_stat)){
  querytrait_ss <- cophe_multi_trait_data$summ_stat[[trait_idx]]
  # Here LD is the same
  querytrait_ss$LD <- LD
  trait_variant_pair <- paste0('Trait', trait_idx, '_', querysnpid)
  res.multi.lbf[[trait_variant_pair]] <- cophe.susie.lbf(querytrait_ss, querysnpid = querysnpid, querytrait = paste0('Trait', trait_idx))
}

res.multi.lbf.df = bind_rows(res.multi.lbf)

```


```{r}
head(res.multi.lbf.df)
```

**Note: ** The output of `cophe.susie` or `cophe.single` can also be used as input to the hierarchical model as it has all the fields required for the input. This would be useful when you would like to compare results from the fixed priors to those obtained from priors inferred using the hierarchical model. [Swap `cophe.susie` for `cophe.susie.lbf` above and instead of bind_rows do : `res.multi.lbf.df = multitrait.simplify(res.multi.lbf)`]

#### Run hierarchical model for priors

The input df for the multi.dat arguments should contain the following fields: "lBF.Ha","lBF.Hc" and "nsnps".  
```{r warning=F, message=F, fig.width = 8, fig.height=8, out.width = "75%",  fig.ncol = 3, fig.align = "center"}
# covar=F
## Set covar to TRUE to include covariates
covar=T
covar_vec = cophe_multi_trait_data$covar_vec
cophe.hier.res <- run_metrop_priors(res.multi.lbf.df, avg_posterior=T, avg_pik = T, covar_vec = covar_vec, covar = covar, nits = 30000)
names(cophe.hier.res)
```
Note: Setting posterior or pik to TRUE is memory intensive for very large datasets

#### Diagnostics of the hierarchical model
```{r mcmcDiagnostics, warning=F, message=F, fig.width = 6, fig.height=6, out.width = "50%", fig.align = "center"}
loglik <- cophe.hier.res$ll
parameters <- cophe.hier.res$parameters
col <- rgb(red = 0.4, green = 0.7, blue = 0.5, alpha = 0.8)
par(mfrow=c(2,2))
plot(1:length(loglik), loglik, main="loglik",type="l", col=col, ylab = "ll", xlab="")
plot(1:ncol(parameters), parameters[1,], main="alpha",type="l", col=col, ylab = "alpha", xlab="")
plot(1:ncol(parameters), parameters[2,], main="beta",type="l", col=col, ylab = "beta", xlab="")
if (covar == T)
  plot(1:ncol(parameters), parameters[3,], main="gamma",type="l", col=col, ylab = "gamma", xlab="")

```






#### Prediction

`cophe.hier.res$avg.posterior` contains the posterior probabilities of the hypotheses : $H_n$,  $H_a$ and $H_c$ for the queryvariant/querytraut pairs obtained from the hierarchical model.
```{r}
res.post.prob = cbind(cophe.hier.res$avg.posterior, cophe.hier.res$data)
```

```{r}
res.hier.predict <- cophe.hyp.predict(as.data.frame(res.post.prob ))
knitr::kable(res.post.prob, row.names = F, digits=3)
```


#### Visualisation

Use the `cophe_plot` function to return -log10(pval), ppHa and ppHc PheWAS plots from the cophescan output

```{r cophePlots, warning=F, message=F, fig.width = 9, fig.height=6, out.width = "100%", fig.align = "center"}
res.plots = cophe_plot(res.hier.predict, traits.dat = cophe_multi_trait_data$summ_stat, querysnpid = querysnpid, pheno_names = paste0('Trait', 1:30))

ggpubr::ggarrange(res.plots$pval, res.plots$ppHa, res.plots$ppHc, ncol = 2, nrow = 2)
```



